---
title: "Getting Started with SurplusProductionModel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with SurplusProductionModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **SurplusProductionModel** package fits Pella-Tomlinson surplus production
models to fishery catch and abundance index (CPUE) data using maximum likelihood
estimation. 

The Pella-Tomlinson model generalises the classic Schaefer model by adding a
shape parameter *m* that controls the location of maximum sustainable yield
(MSY) relative to carrying capacity *K*.

Key capabilities:
- Fit single-area or multi-area models
- Accommodate multiple CPUE indices (e.g. longline and trawl survey)
- Fix or constrain parameters for improved identifiability
- Calculate biological reference points (MSY, B~MSY~, F~MSY~)
- Diagnostic plotting of biomass trajectories, CPUE fits, and residuals

## Installation

```r
# Install the development version from GitHub
# install.packages("devtools")
devtools::install_github("alistairdunn1/SurplusProductionModel")
```

## Example Data

The package ships with simulated data loosely based on the Ross Sea Antarctic
toothfish fishery. These are generated from known parameters, so you can
compare the model estimates with the truth.

```{r load-package}
library(SurplusProductionModel)
```

```{r load-data}
data(ross_sea_catch)
data(ross_sea_cpue_simple)
data(ross_sea_true_params)
data(ross_sea_true_biomass)
```

The catch data cover 1997--2024 and contain annual catches in tonnes. The
trajectory includes a development phase, peak exploitation, and a catch
reduction period that allows partial stock recovery:

```{r show-catch}
head(ross_sea_catch)
```

The single-index CPUE data provide a longline abundance index from 1998
onward:

```{r show-cpue}
head(ross_sea_cpue_simple)
```

Let's visualise the input data:

```{r plot-inputs, fig.height = 4}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(ross_sea_catch$year, ross_sea_catch$catch, type = "b", pch = 16,
     xlab = "Year", ylab = "Catch (t)", main = "Catch History")
plot(ross_sea_cpue_simple$year, ross_sea_cpue_simple$cpue, type = "b", pch = 16,
     xlab = "Year", ylab = "CPUE (kg/hook)", main = "Longline CPUE")
```

## Fitting a Model

Prepare the data as a list with `cpue_data` and `catch_data` elements, then
call `fit_pella_tomlinson_model()`.

In surplus production models, the shape parameter *m* is typically poorly
identified from CPUE data alone. Standard practice is to fix *m* (e.g.
*m* = 2 for the classic Schaefer model) using the `fixed_params` option:

```{r fit-model, warning = FALSE}
model_data <- list(
  cpue_data  = ross_sea_cpue_simple,
  catch_data = ross_sea_catch
)

fit <- fit_pella_tomlinson_model(
  data    = model_data,
  options = list(
    fixed_params = list(log_m = log(2)),
    control = list(eval.max = 5000, iter.max = 2000)
  )
)
```

The returned object is an S4 `ProductionModel` that summarises fit quality
and estimated parameters:

```{r print-model}
print(fit)
```

The estimated natural-scale parameters:

```{r show-params}
fitted_params <- fit$results$parameters
# Single-area fits store per-area names (e.g. "K.A1", "q.A1"); accept either.
getp <- function(nm) unname(fitted_params[startsWith(names(fitted_params), nm)][1])
cat(sprintf("  r         = %.4f  (intrinsic growth rate)\n", getp("r")))
cat(sprintf("  K         = %.0f  (carrying capacity, t)\n",  getp("K")))
cat(sprintf("  m         = %.1f    (shape, fixed)\n",        getp("m")))
cat(sprintf("  sigma_obs = %.3f  (observation error SD)\n",  getp("sigma_obs")))
cat(sprintf("  q         = %.6f  (catchability)\n",          getp("q")))
cat(sprintf("  d0        = %.3f  (initial depletion)\n",     getp("d0")))
cat(sprintf("  B_initial = %.0f  (initial biomass = d0 * K, t)\n", getp("d0") * getp("K")))
```

## Reference Points

Calculate biological reference points from the fitted model with
`calculate_reference_points()`:

```{r ref-points}
rp <- calculate_reference_points(fit)
print(rp)
```

## Comparing to Truth

Because this dataset was generated with known parameters, we can compare the
model estimates with the true values stored in `ross_sea_true_params`:

```{r compare-params}
true <- ross_sea_true_params

cat(sprintf("Parameter  %10s  %10s\n", "True", "Estimated"))
cat(sprintf("r          %10.4f  %10.4f\n", true$r,    getp("r")))
cat(sprintf("K          %10.0f  %10.0f\n", true$K,    getp("K")))
cat(sprintf("m          %10.2f  %10.2f\n", true$m,    getp("m")))
cat(sprintf("q          %10.6f  %10.6f\n", true$q_LL, getp("q")))
cat(sprintf("B_initial  %10.0f  %10.0f\n", true$B0,   getp("d0") * getp("K")))
```

Note that exact recovery of the generating parameters is not expected from
maximum likelihood estimation with a finite, noisy dataset. In particular,
surplus production models are known to have ridges in the likelihood surface
where different combinations of *r*, *K*, and *B*~0~ produce similar fits.
Bayesian methods with informative priors are commonly used to address this
identifiability issue in practice.

We can overlay the estimated biomass trajectory on the true trajectory:

```{r compare-biomass, fig.height = 4}
est_biomass <- fit$results$biomass
years <- fit$data$years

plot(ross_sea_true_biomass$year, ross_sea_true_biomass$biomass,
     type = "l", lwd = 2, col = "grey40",
     xlab = "Year", ylab = "Biomass (t)",
     main = "Estimated vs True Biomass",
     ylim = range(c(ross_sea_true_biomass$biomass, est_biomass), na.rm = TRUE))
lines(years, est_biomass, col = "#0072B2", lwd = 2, lty = 2)
legend("topright", legend = c("True", "Estimated"),
       col = c("grey40", "#0072B2"), lwd = 2, lty = c(1, 2))
```

## Diagnostic Plots

The `plot_model_fit()` function produces four diagnostic panels: biomass
trajectory, observed vs fitted CPUE, residuals, and harvest rate.

```{r diagnostics, fig.height = 8, fig.width = 7, results = "hide", message = FALSE}
plots <- plot_model_fit(fit, which = "all")
```

## Fitting with Multiple CPUE Indices

The package also supports multiple abundance indices. The `ross_sea_cpue`
dataset contains both longline and trawl survey CPUE:

```{r multi-index}
data(ross_sea_cpue)
head(ross_sea_cpue)
table(ross_sea_cpue$label)
```

To use both indices, simply pass the multi-index data frame. Each index
receives its own estimated catchability *q*, while sharing the underlying
population dynamics parameters.

```{r fit-multi, eval = FALSE}
multi_data <- list(
  cpue_data  = ross_sea_cpue,
  catch_data = ross_sea_catch
)

fit_multi <- fit_pella_tomlinson_model(
  data    = multi_data,
  options = list(
    fixed_params = list(log_m = log(2)),
    control = list(eval.max = 5000, iter.max = 2000)
  )
)

print(fit_multi)
```

## Model Formulation

The Pella-Tomlinson surplus production model used in this package is:

**Production function (standard Pella-Tomlinson):**

$$P(B) = \frac{r \, B}{m - 1} \left[ 1 - \left(\frac{B}{K}\right)^{m-1} \right]$$

Here *r* is the intrinsic growth rate. The model reduces to the Schaefer
form $P(B) = r\,B\,(1 - B/K)$ at $m = 2$ and, in the limit $m \to 1$, to the
Fox form $P(B) = r\,B\,\log(K/B)$. The maximum-sustainable-yield reference
points are $B_{\text{MSY}} = K\,m^{-1/(m-1)}$, $F_{\text{MSY}} = r/m$, and
$\text{MSY} = F_{\text{MSY}}\,B_{\text{MSY}}$.

**State equation (biomass dynamics):**

$$B_{t+1} = B_t + P(B_t) - C_t + \varepsilon_t, \quad \varepsilon_t \sim N(0, \sigma^2_{\text{proc}})$$

**Observation equation:**

$$\text{CPUE}_t = q \, B_t \, e^{\eta_t}, \quad \eta_t \sim N(0, \sigma^2_{\text{obs}})$$

Special cases:

- **Schaefer model** ($m = 2$): $B_{\text{MSY}} = K/2$
- **Fox model** ($m \to 1$): $B_{\text{MSY}} = K/e$

## Next Steps

- Explore multi-area spatial models by adding an `area` column to your data
- Examine sensitivity to the shape parameter *m*
- Compare AIC/BIC across models with different structures
- Use `validate_cpue_data()` and `validate_catch_data()` to check data quality
  before fitting

```{r session-info}
sessionInfo()
```
